---
title: "Predicting Breast Cancer using Logistic Regression, Decision Trees, and Naïve Bayes"
author: "Hugo Argueta, Karina Pizano, Yarely Chino"
date: "May 12, 2017"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, comment="", echo=TRUE)
```

```{r echo=FALSE}
library(rpart)
library(rpart.plot)

library(e1071)
library(rpart)
# source("/Users/karinapizano/Documents/Data Mining/lin-regr-util.R")
```

## Introduction

#### _We wanted to predict the presence of breast cancer with an accuracy of at least 85%, based on features that describe characteristics of cancer cell nuclei present in an image._

### Load data

```{r}
set.seed(135)

dat = read.csv("/Users/juice/Documents/CST463/final-project/breast-cancer-prediction/data.csv")
dat$X = NULL
dat$output = ifelse(dat$diagnosis == "M", 1, 0)

# Create Factors
dat$diagnosis = factor(dat$diagnosis, labels =c("None", "Cancer"))
```

### Split dataset into training set and testing set

```{r}
num_tr = sample(1:nrow(dat), floor(nrow(dat) * .75))
tr_dat = dat[num_tr,]
te_dat = dat[-num_tr,]
```

### Data Exploration

_Scatter plots to detect any correlation between the features. We see strong positive trends between radius mean and perimeter mean, as well as perimeter mean and concavity mean._

```{r}
plot(dat[,c("radius_mean", "texture_mean","perimeter_mean", "concavity_mean")])
```

_Our dataset contains a good balance of diagnosis data._
```{r}
barplot(table(dat$diagnosis), names.arg=c("No Cancer", "Cancer"), main="Amount of Cancer and Non Cancer Patients", col=c("blue", "red4"))
```

_Function to predict breast cancer using a single element._
```{r}
predict_by = function(name){
  plot(dat$output ~ dat$element, col="red4", pch = 16, ylab = "probability", xlab = name, main=paste0("Breast cancer probability using ", name))
  fit = glm(output ~ element, data=dat, family=binomial )
  bp = data.frame(element=seq(range(dat$element)[1],range(dat$element)[2], length.out=50))
  probs = predict(fit, newdata=bp, type="response")
  lines(bp$element, probs, lty=2, col = "blue")
}
```

####Logistic function using radius worst
```{r}
dat$element = dat$radius_worst
predict_by("radius worst")

```


####Logistic function using area standard error
```{r warning=FALSE}
dat$element = dat$area_se
predict_by("area standard error")

```

_Here, we are plotting texture worst by radius worst and displaying the occurrences of breast cancer in red, and the occurrences of no breast cancer in green, for each point in our plot. This plot shows that there is a boundary line at around 16. This indicates that these two features will be useful in predicting breast cancer._

```{r}
plot(dat$texture_worst, dat$radius_worst, col = "green", main = "Texture Worst by Radius Worst", xlab = "texture worst", ylab = "radius worst", pch=16)
malignant = dat$diagnosis == "Cancer"
points(dat[malignant, ]$texture_worst, dat[malignant, ]$radius_worst, col = "red", pch=16)
text_rad_fit = rpart(diagnosis ~ texture_worst + radius_worst, data=tr_dat, method="class")
split = text_rad_fit$splits[1, "index"]
abline(h=split, lty=2, col='blue')
```

_Density plot for mean and standard deviation of concave points worst. Here we can see that when the feature concave points reaches .13 their is a clear division between having cancer and not having cancer._
```{r}
plot(density(dat[dat$diagnosis == "None",]$concave.points_worst), col="green", main="Mean")
lines(density(dat[dat$diagnosis == "Cancer",]$concave.points_worst), col="red")
legend(-.02,11,legend =c("Cancer", "No Cancer"), fill=c("red", "green"), cex=0.75)

```

### Model Creation

#### Creating Logistic Regression

_We are creating a logistic regression model using radius worst, area worst, and smoothness worst in order to predict breast cancer._

```{r warning=FALSE}
logistic_fit = glm(output ~ radius_worst + area_worst + smoothness_worst, dat=tr_dat, family=binomial)

```

#### Creating Decision Tree Classifier

_We are creating a decision tree classifier to predict the diagnosis feature based on the features radius worst, concave.points worst, and texture worst._

```{r}
tree_fit = rpart(diagnosis ~ radius_worst + concave.points_worst + texture_worst, data=tr_dat, method="class")

prp(tree_fit, extra=106, varlen=-10, main="Classification Tree for Breast Cancer", box.col=c("green", "red")[tree_fit$frame$yval])
```

#### Creating Naïve Bayes Classifier

_We are creating a Naïve Bayes classifier to predict the diagnosis feature based on the features radius worst, texture worst, and concave.points worst._

```{r }
bayes_fit = naiveBayes(diagnosis ~ radius_worst + texture_worst + concave.points_worst, data= tr_dat)
```

### Assessing the Models

#### Confusion matrix for Logistic Regression

_From the confusion matrix for actuals vs. predicted, we can see that we have low false positives and low false negatives._
```{r}
pred = predict(logistic_fit, newdata=te_dat, type="response")
predicts = as.numeric(pred > 0.5)
actuals = te_dat$output
conf_mtx = table(predicts, actuals)
row.names(conf_mtx) = c("B", "M")
colnames(conf_mtx) = c("B", "M")
conf_mtx

```

#### Confusion matrix for Decision Tree
_From the confusion matrix for actuals vs. predicted, we can see that we have low false positives and no false negatives._

```{r}
tree_predicts = predict(tree_fit, te_dat, type="class")
tree_actuals = te_dat$diagnosis
tree_conf_mtx = table(tree_actuals, tree_predicts)
row.names(tree_conf_mtx) = c("B", "M")
tree_conf_mtx
```

#### Confusion matrix for Naïve Bayes
_From the confusion matrix for actuals vs. predicted, we can see that we also have low false positives and low false negatives._

```{r}
bayes_predicts = predict(bayes_fit, te_dat)
bayes_actuals = te_dat$diagnosis

bayes_conf_mtx = table(bayes_actuals, bayes_predicts)
row.names(bayes_conf_mtx) = c("B", "M")
bayes_conf_mtx
```

###Accuracy of our Classifiers

_Logistic Regression_
```{r}
round(mean(predicts == actuals), 3)
```

_Tree Classifier_
```{r}
round(mean(tree_predicts == tree_actuals), 3)
```

_Naive Bayes Classifier_
```{r}
accuracy = round(mean(bayes_predicts == te_dat$diagnosis), 3)
accuracy
```

### Precision of our Classifiers

_Logistic Regression_
```{r}
conf_mtx[2,2] / (conf_mtx[2,2] + conf_mtx[1,2])
```

_Tree Classification_
```{r}
tree_conf_mtx[2,2] / (tree_conf_mtx[2,2] + tree_conf_mtx[1,2])
```

_Naive Bayes_
```{r}
bayes_conf_mtx[2,2] / (bayes_conf_mtx[2,2] + bayes_conf_mtx[1,2])
```

### Recall of our Classifier

_Logistic Regression_
```{r}
conf_mtx[2,2] / (conf_mtx[2,2] + conf_mtx[2,1])
```

_Tree Classification_
```{r}
tree_conf_mtx[2,2] / (tree_conf_mtx[2,2] + tree_conf_mtx[2,1])
```

_Naive Bayes_
```{r}
bayes_conf_mtx[2,2] / (bayes_conf_mtx[2,2] + bayes_conf_mtx[2,1])
```

### Learning Curves for Classifiers

#### Logistic Regression

_From this learning curve, we can see that this classifier has high variance getting more training data will help the classifier._

```{r echo=FALSE, warning=FALSE}
te_errs = c()
tr_errs = c()
te_actual = te_dat$output

tr_sizes = seq(100, nrow(tr_dat), length.out=10)

for (tr_size in tr_sizes) {
  tr_dat1 = tr_dat[1:tr_size,]
  tr_actual = tr_dat1$output
  
  #creating model using tr_dat1
  fit = glm(output ~ radius_worst + area_worst + smoothness_worst, dat=tr_dat1, family=binomial)
  
  # error on training set
  tr_predicted = predict(fit, tr_dat1, type="response")
  tr_predicted = as.numeric(tr_predicted > 0.5)
  err = sum(tr_actual != tr_predicted)/length(tr_predicted)
  tr_errs = c(tr_errs, err)
  
  # error on test set
  te_predicted = predict(fit, te_dat, type="response")
  te_predicted = as.numeric(te_predicted > 0.5)
  err = sum(te_actual != te_predicted)/length(te_predicted)
  te_errs = c(te_errs, err)
}
 plot(tr_sizes, te_errs, type="l", col="red4", ylim=range(c(te_errs,tr_errs)), xlab="Size", ylab="Error")
 lines(tr_sizes, tr_errs, col="blue")
legend(range(c(te_errs,tr_errs))[2], x = 350, legend = c("training errors", "testing errors"), col = c("blue", "red4"), pch="|",  cex = .8)
```

#### Decision Tree Classification

_From this learning curve, we can see that this classifier has low bias and low variance because the training errors and the test error lines converge as the training set size increases._

```{r echo=FALSE}
te_errs = c()
tr_errs = c()
te_actual = te_dat$diagnosis
tr_sizes = seq(100, nrow(tr_dat), length.out=10)
for (tr_size in tr_sizes) {
  tr_dat1 = tr_dat[1:tr_size,]
  tr_actual = tr_dat1$diagnosis
  fit = rpart(diagnosis ~ radius_worst + concave.points_worst + texture_worst, data=tr_dat1, method="class")
  # error on training set
  tr_predicted = predict(fit, tr_dat1, type="class")
  tr_predicted
  err = sum(tr_actual != tr_predicted)/length(tr_predicted)
  tr_errs = c(tr_errs, err)
 
  # error on test set
  te_predicted = predict(fit, te_dat, type="class")
  err = sum(te_actual != te_predicted)/length(te_predicted)
  te_errs = c(te_errs, err)
  te_errs
}

plot(range(tr_sizes), range(te_errs, tr_errs), type='n', xlab='training set size', ylab='error')

lines(tr_sizes, tr_errs, col='green')
lines(tr_sizes, te_errs, col='red4')

legend(range(c(te_errs,tr_errs))[2], x = 350, legend = c("training errors", "testing errors"), col = c("green", "red4"), pch="|",  cex = .8)
```

#### Naive Bayes Classifier

_The learning curves shows, that we have high variance. We need more data to train and test. Overall the errors decrease for both the test and training data._

```{r echo=FALSE}

te_errs = c()
tr_errs = c()
te_actual = te_dat$diagnosis
tr_sizes = seq(100, nrow(tr_dat), length.out=10)

for (tr_size in tr_sizes) {
  tr_dat1 = tr_dat[1:tr_size,]
  tr_actual = tr_dat1$diagnosis
  fit = naiveBayes(diagnosis ~ radius_worst + texture_worst + concave.points_worst, data=tr_dat1)
  
  # error on training set
  tr_predicted = predict(fit, tr_dat1, type="class")
  tr_predicted
  err = sum(tr_actual != tr_predicted)/length(tr_predicted)
  tr_errs = c(tr_errs, err)
 
  # error on test set
  te_predicted = predict(fit, te_dat, type="class")
  err = sum(te_actual != te_predicted)/length(te_predicted)
  te_errs = c(te_errs, err)
  te_errs
}

plot(range(tr_sizes), range(te_errs, tr_errs), type='n', xlab='training set size', ylab='error')

lines(tr_sizes, tr_errs, col='purple')
lines(tr_sizes, te_errs, col='red4')

legend(range(c(te_errs,tr_errs))[2], x = 350, legend = c("training errors", "testing errors"), col = c("purple", "red4"), pch="|",  cex = .8)
```


### Additional Logistic Regression Model Assessments

_Function that takes in predicts and actuals to produce a dataframe that contains model-assessing information, using different thresholds._
```{r}
prec_recall_summary = function(predicts, actuals) {
   thresh = seq(0, 1, length.out=50)
   prec_rec = data.frame()
   actuals = factor(as.numeric(actuals))
   for (th in thresh) {
     predicts = factor(as.numeric(pred >= th), levels=c("0","1"))
     prec_rec = rbind(prec_rec, as.vector(table(predicts, actuals)))
   }
   names(prec_rec) = c("TN", "FP", "FN", "TP")
   prec_rec$threshold = thresh
   prec_rec$precision = prec_rec$TP/(prec_rec$TP + prec_rec$FP)
   prec_rec$recall    = prec_rec$TP/(prec_rec$TP + prec_rec$FN)
   prec_rec$false_pos = prec_rec$FP/(prec_rec$FP + prec_rec$TN)
   return(prec_rec)
 }
 prec_rec = prec_recall_summary(predicts, actuals)

```



####Plots show precision and recall for Logistic Regression

```{r}
par(mfrow=c(2,1))
plot(prec_rec$threshold, prec_rec$precision, col = "red4", type = "l", xlab = "threshold", ylab = "precision" )
grid(col="grey60")
plot(prec_rec$threshold, prec_rec$recall, col = "red4", type = "l", xlab = "threshold", ylab = "recall" )
grid(col="grey60")

```

####Receiver Operating Character Graph for Logistic Regrassion

```{r}

prec_rec$true_pos = prec_rec$TP / (prec_rec$TP + prec_rec$FN)
plot( prec_rec$false_pos, prec_rec$true_pos, col = "red4", type = "l", xlab = "True positive rate", ylab = "False positive rate" , main= "ROC")
grid(col="grey60")
```

## Conclusion
_In conclusion, we were able to obtain an average of 96.6% accuracy when using Logistic Regression, Decision Trees, and Naïve Byes. The classifier that achieved the highest accuracy was _

