---
title: "finalProject-breastCancer"
author: "Karina Pizano"
date: "5/8/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(e1071)
source("/Users/karinapizano/Documents/Data Mining/lin-regr-util.R")
```


<!-- Title -->
###Breast Cancer Prediction
_Data Cleaning and Munging_
```{r }
dat = read.csv("data.csv")

# Create Factors
dat$diagnosis = factor(dat$diagnosis, labels =c("None", "Cancer"))

# Part the data.
data_sets = split_data(dat)
tr_dat = data_sets[[1]]
te_dat = data_sets[[2]]

# Remove bad columns and data.
dat$X = NULL
```

##Data Exploration
_Scatter Plot to detect any correlation between the attributes._
<!-- Scatter plots to detect any correlation, if any -->
```{r}
plot(dat[,c("radius_mean", "texture_mean","perimeter_mean", "concavity_mean")])
```

<!-- Plot all the features, to see if there is any correlation -->
_We can see more correlation between more attributes_
```{r}
plot(dat[,1:10])
```


<!-- Our Data -->
_Our data contains a good amount of data for both having Cancer and not having Caner, which will make our predictions pretty accurate._
```{r}
barplot(table(dat$diagnosis), names.arg=c("No Cancer", "Cancer"), main="Amount of Cancer and Non Cancer patients", col=c("green", "yellow"))
```


<!-- Hist of radius_mean -->

_Whats a popular radius mean when looking an infected area_

```{r}
hist(dat$radius_mean, main="Radius mean", col="red", xlab = "radius mean")
```


##Naive Bayes Model
<!-- Naive Bayes Model -->
_Create a Naive Bayes Model_
```{r }
fit = naiveBayes(diagnosis ~ radius_worst + texture_worst + concave.points_worst, data= tr_dat)
```

<!--Summary of fit -->
_Summary of the Model_
```{r }
summary(fit)
```

<!-- Generate Predictions -->
_Predictions & Confusion Matrix_
```{r}
predictions = predict(fit, newdata=te_dat)
con_matrix = table(te_dat$diagnosis, predictions)
con_matrix
```


<!--Precision -->
_Precision of the Classifier_
```{r}
con_matrix[2,2] / (con_matrix[2,2] + con_matrix[1,2])
```



<!-- Recall -->
_Recall of our Classifier_
```{r}
con_matrix[2,2] / (con_matrix[2,2] + con_matrix[2,1])
```




<!-- Accuracy of the Classifier -->
_The accuracy of our classifier._
```{r}
accuracy = mean(predictions == te_dat$diagnosis)
accuracy
```

<!-- Density Plot "RADIUS_WORST" -->
_Showing the distribution of radius worst when a person is diagnosed with cancer or not._
```{r}

plot(density(dat[dat$diagnosis == "None",]$radius_worst), col="green", main="Distibution of radius_mean")
lines(density(dat[dat$diagnosis == "Cancer",]$radius_worst), col="red")
legend(7,0.18,legend =c("Cancer", "No Cancer"), fill=c("red", "green"), cex=0.75)

```
<!-- Fitted model Standard deviations and mean using radius_worst when the disease is present and not present -->
_We are looking at the trained distributions for the fitted model. Mean and Standard Deviation._
```{r}
fit$tables$radius_worst
```




<!-- Density plot mean and standard deviation using "RADIUS_WORST" -->
_Density plot for mean and standard deviation of radius worst._
```{r}
plot(density(fit$tables$radius_worst[1,]), col="red", main="Mean & Standard Deviation")
lines(density(fit$tables$radius_worst[2,]), col="green")

```



<!-- Density plot "WORST TEXTURE" -->
_Density plot for mean and standard deviation of worst texture._
```{r}
plot(density(dat[dat$diagnosis == "None",]$texture_worst), col="green", main="Mean")
lines(density(dat[dat$diagnosis == "Cancer",]$texture_worst), col="red")
legend(50,0.002,legend =c("Cancer", "No Cancer"), fill=c("red", "green"), cex=0.75)
```










